\begin{abstract}
Modern end-to-end encrypted (E2EE) messaging applications offer a wide array of \emph{message deletion} features, such as unsending messages, closing conversations, view-once media, and more. 
These features provide an extra layer of privacy, 
% beyond in-transit encryption
allowing users to hide sensitive messages even from adversaries who may gain access to their devices. Privacy of deleted messages can be critical in many settings, such as compelled device searches, known-adversary threat models, and more. Yet, despite its ubiquity, message deletion remains surprisingly understudied: we have little understanding of the security offered by such features in practice, let alone formal definitions of what their security \emph{should} mean.

This work initiates the formal study of \emph{message deletion security} in E2EE applications. 
We first investigate the security of message deletion in practice, using two leading E2EE applications as case studies: WhatsApp and Signal. Our analysis reveals an unsatisfactory landscape: we identify numerous security issues that enable recovery of deleted messages from a victim's client-side state, across multiple message types (text, attachments, and more), and affecting all their deletion features. These vulnerabilities stem from general design patterns, common to both applications, and thus lead to higher-level lessons for E2EE application design.

Our findings motivate that secure deletion poses a challenge for E2EE applications, and so foundations for its principled study are critical. We thus provide the first formal treatment of client-side deletion security for E2EE applications. We introduce syntax that models the core components of E2EE clients relevant to deletion, and define security notions that formalize our privacy goals. To exercise our security model, we then analyze both the vulnerabilities we identified and the mitigations we propose, showing that our work provides solid groundwork for the study of message deletion as a first-class security principle for E2EE applications. 
\end{abstract}