\section{Message Deletion in the Wild}\label{sec:attacks}
Having introduced our threat model, we now turn to the motivating question behind this work: do real-world applications safely delete messages? We explore this question through the lens of two case-study applications, WhatsApp and Signal. We reverse engineered implementation details of their message deletion features, and performed a security analysis in the context of our threat model. Our analysis reveals an unsatisfactory landscape: we show numerous vulnerabilities in both applications (six in WhatsApp, and seven in Signal), allowing for recovery of deleted messages across all deletion features. 

% As we will see, some of the security issues we find are simple, resulting from improper cleanup of database tables; while others are more subtle, leveraging client-side logic spanning multiple parts of the application state. 


\paragraph{Summary of findings.} Our security analysis is structured around three types of messages, which together encompass the full range of application content: texts, attachments (e.g., media files), and user interactions with messages (e.g., reactions, pinning messages, etc). Our goal is to understand, for each message type, whether it is possible to recover deleted content of this type (and under which deletion features). In both WhatsApp and Signal, we show security issues encompassing all three message types.

Our findings are rooted in three \emph{general} design patterns, corresponding to each of the three message types. For text, the root cause of vulnerabilities is \emph{deletion from search indeces}, and the fact that it is implemented \emph{logically} instead of \emph{physically}. For attachments, the root cause is \emph{file deduplication}, from which one can reverse-engineer whether deleted attachments had some relationship with existing ones. Lastly, for message interactions, the root cause is \emph{straggling metadata} in database tables. Interestingly, these same issues are present in both WhatsApp and Signal. Our findings thus lead to higher-level lessons and takeaways, which we unpack in Section~\ref{sec:discussion}.

We show a summary of our vulnerabilities in Table~\ref{fig:vulnerabilities-summary}. (In addition to the three root issues mentioned above, we also identify and report other bespoke, application-specific security issues.) As we will see, some vulnerabilities require no pre-conditions, while others rely on (realistic) assumptions about the content or timing of messages.

Before we explain these in detail, we first summarize our experimental setup, and provide an overview of the client-side architecture of our case studies.

\paragraph{Methodology and experimental setup.} We performed our analysis of both applications using Android devices, with the latest version at the time of each application installed.\footnote{Version 2.26.11.74 for WhatsApp, and \andres{TODO} for Signal.} (As we discuss later on, however, we believe our findings are general, and not Android-specific.) Similar to prior works on attacks against messaging applications~\cite{fabrega2024injection}, we used emulated devices~\cite{android-emulator-1} for ease of experimentation, allowing us to easily reset the phones if needed and test under numerous setups and conditions. This was sufficient for our purposes, since emulated devices copy both the software \emph{and} hardware of physical devices~\cite{android-emulator-2}, and thus application files (which are the key component of our threat model) are identical as on real devices; indeed, the main use case of emulated devices is for testing applications during development, to be later run on physical end-user devices~\cite{android-emulator-1}.

Recall that deletion security considers an adversary with full, physical access to a device, which allows them to obtain the local application state. In general, access to a device is sufficient for access to its local files (e.g., by performing a full memory dump or device backup). For specific device types, however, there are often easier ways to navigate their file system. For example, in Android, a simple solution is to enable root access~\cite{sun2015android} on the device, which gives unfettered access to its file system. We opt for this approach in our experiments, but note that our forthcoming security analyses are agnostic to the method used to inspect the device's content. In particular, root access allows us to use Android Debug Bridge~\cite{adb} to explore the device's file system via a command-line interface.

With this setup, at a high-level our methodology had a simple recipe: we would send messages from one device to another, record a snapshot of the application files, ... \andres{This required a combination of logical-level changes (e.g., the SQL statements that change) and physical level changes (e.g., serialized binaries)}


\paragraph{Architecture overview.} WhatsApp and Signal have similar high-level structures to their client-side states. We provide a general overview here, and fill in additional details as needed during our vulnerability descriptions that follow. 

As reported by prior work~\cite{fabrega2024exploiting}, both applications store information about sent and received messages in a single SQLite database, which is serialized using SQLite's standard database file format~\cite{sqlite-database-file-format}. The schemas for these databases are complex, involving (at the time of writing) 285 tables for WhatsApp and 60 tables for Signal.


Both databases have a main table that acts as the central hub for all sent and received messages, where each row corresponds to a single (sent or received) message, storing its text content, timestamp, chat it is part of, and more. This table, called ``$\messagesTable$'' in both applications, is then referenced by numerous other tables that store more specific message metadata. (We discuss details of $\messagesTable$ and other relevant tables later in the section.) For convenience, moving forward we use shorthand $\databaseWhatsApp$ and $\databaseSignal$ to refer to the database of WhatsApp and Signal, respectively.

For messages that contain attachments (e.g., documents, images, videos, etc.), WhatsApp and Signal store these files as separate binaries in a dedicated BLOB store. Metadata about these files is then stored in the $\mediaTable$ table in $\databaseWhatsApp$ and $\attachmentTable$ table in $\databaseSignal$. In both, each row stores information about a single file, including the primary key in $\messagesTable$ for its corresponding message, the path to the file's location in storage, its size, hash, etc. Similar to the $\messagesTable$ table, $\mediaTable$ (in WhatsApp) and $\attachmentTable$ (in Signal) act as the central hub for media information, and is referenced by auxiliary tables that store more fine-grained media metadata.

In addition to the above, the application state contains numerous other files. However, these are not relevant to our purposes, so we omit description of them for brevity.

We reversed-engineered how this application state (and, in particular databases $\databaseWhatsApp$ and $\databaseSignal$) are updated as a result of WhatsApp and Signal's various message deletion features, following the general methodology mentioned at the above. We now proceed to the results of our security analysis, ordered in terms of the three aforementioned message types: text, attachments, and message reactions.

\subsection{Text Recovery}\label{subsec:attacks-text}
WhatsApp and Signal support four main ways to delete text: chat deletions, local message deletions, unsends, and disappearing messages. (See Section~\ref{sec:e2ee-deletion-systematization} for more details on these.) In this section, we show vulnerabilities against both applications based on \emph{deletion from client-side search indeces}, which allows an adversary to recover the text of deleted messages across all message deletion features via inspection of the serialized index. This vulnerability is fairly technical, relying on many low-level details of search indeces and how they are encoded in databases. So, due to space constraints, we highlight the main ideas here, and defer full details to Appendix~\ref{}. We discuss each application in turn.

\paragraph{WhatsApp's search index.}
To support efficient search, WhatsApp uses SQLite's FTS4 module~\cite{sqlite-fts4} to implement full-text search on all sent and received messages. An FTS4 index is logically structured as an \emph{inverted} index, mapping \emph{terms} to the \emph{documents} (and their positions therein) in which they appear. In WhatsApp, terms (i.e., index entries) roughly correspond to individual words,\footnote{More precisely, in WhatsApp, the terms in message $m$ are the words that result after casting into $m$ all lowercase, and replacing all non-alphanumeric characters with whitespace.} and documents to the messages in which they are contained; messages are represented by a \emph{message identifier}, which are assigned sequentially (and are also the primary keys in the $\messagesTable$ table). Thus, the index has an individual entry for every word across all messages, mapping to a list of all messages (and their position within the message) in which the word is present. For example, if a user has (in order) messages ``Hi Alice'' and ``Hi Bob'', their index would map ``hi'' to $[(1, 1), (2, 1)]$, ``alice'' to $[(1, 2)]$, and ``bob'' to $[(2, 2)]$.

Like all other message information, the index is stored in database $\databaseWhatsApp$. The way in which an FTS4 index is encoded and arranged within a database is complex; for clarity, we provide a simplified treatment here, and refer readers to 
Appendix~\ref{} and~\cite{sqlite-fts4} for details.
At a high-level, an FTS4 index is organized as a collection of multiple B-trees, with term-document mappings (i.e., word-message mappings in WhatsApp) stored in the leaves. Concretely, a leaf contains a sequence of term and \emph{document list} (doclist) pairs, where a doclist consists of document-position pairs (e.g., $[(1, 1), (2, 1)]$ in our example above). Term-doclist pairs within a leaf are ordered lexicographically by term, and document-position pairs within a doclist are ordered by document id.
See Figure~\ref{} for an example. 

The B-trees are organized into levels. Whenever a message is added to the index, a new level-0 B-tree is created, containing only the terms in this message, each mapping to the message's identifier (i.e., its primary key in $\messagesTable$) and its position in it. That is, let $m = (w_1, \dots, w_q)$ be the $n$-th (sent or received) message, where $w_i$ are its terms, in order. This results in a level-0 B-tree containing (only) pairs $(w_1, [(n, 1)]), (w_2, [(n, 2)]), \dots, (w_q, [(n, q)])$, arranged lexicographically by $w_i$.\footnote{If there are two $w_i$ and $w_j$ such that $w_i = w_j$ and $i < j$, then there is only a single pair for them of the form $(w_i, [(n, i), [n, j]])$.} After a certain number of level-0 B-trees are present (16, in the case of WhatsApp), they are merged into a level-1 B-tree, containing the term-doclist pairs of all 16 level-0 B-trees, which are then deleted. In particular, if a term is present in multiple level-0 B-trees, it is stored as single entry in the level-1 B-tree, with all document-position pairs in the term's (single) doclist.\andres{Previous sentence candidate to remove.} This progression continues: in general, after 16 level $i$ B-trees are created, they are replaced by a new level $i+1$ B-tree. So, as the index grows, it consists of multiple B-trees of different levels and sizes, which are traversed during a search query. This means that a term, while unique within a B-tree, may be repeated across different trees, which get consolidated when/if these trees merge.\andres{Previous sentence candidate to remove.}

The nodes of all B-trees are stored in the $\messagesIndexSegdirTable$ and $\messagesIndexSegmentsTable$ tables of $\databaseWhatsApp$---root notes in the former, and all other nodes in the latter. Thus, scanning these tables allows inspecting the physical encoding of the index. The maximum size of a node is the database's page size (4096 bytes in WhatsApp), and so lower-level B-trees often require just a single root node.

\paragraph{Text deletion.}
Application-level deletions trigger a cascade of changes in $\databaseWhatsApp$. These changes are generally consistent across all message deletion features. First, the message's text is removed from the $\messagesTable$ table. For silent features (property \#3 in our systematization) such as local deletion, the row is deleted entirely; and for visible features such as unsending messages, the row is kept, but the cell storing its text content is emptied, replaced by a NULL value. Notice that, for silent features, the $\messagesTable$ table thus leaks some information: row deletions result in a \emph{skip} in primary keys, revealing that \emph{some} message was deleted (but not which or in which chat). That is, if the $i$-th message is deleted, the table will have a row with primary key $i-1$ followed by one with $i+1$.

While identifying that deletion occurred can be sensitive in many contexts, our focus is on message recovery. In addition to the $\messagesTable$ table, the second prominent change is that deletion features trigger a deletion request to the index. However, deletions in FTS4 are only \emph{logical}: instead of \emph{erasing} content, deletions create new \emph{tombstone entries}. Concretely, deleting the $n$-th message $m = (w_1, \dots, w_q)$ creates a \emph{new} index entry containing $(w_1, [(n, 0)]), (w_2, [(n, 0)]), \dots, (w_q, [(n, 0)])$. As usual, this entry is stored as a new level-0 B-tree, stored in $\messagesIndexSegdirTable$ (and $\messagesIndexSegmentsTable$, if the message is long and does not fit in a single root node). Such an entry encodes that terms $w_i$ appear in ``position 0'' of message $n$, which instructs future search queries to omit $m$ from the search results for any $w_i$.

Deletions in the search index are thus only at the application-level: both the original index entry (when the message was sent or received) and the new tombstone entry are still present in the physical index in storage, and can therefore be recovered via direct inspection of the B-tree nodes in the $\messagesIndexSegdirTable$ and $\messagesIndexSegmentsTable$ tables.
We show an example in Figure~\ref{}, where we show how the index evolves as a result of message delivery and deletion.

Original entries and tombstone entries for a deleted message are only combined when their respective B-trees merge, in which case tombstone entries overwrite original entries. This requires the victim to receive enough messages to trigger a merge, where the exact number depends on how ``far apart'' both index entries are. Finally, the tombstone entry is removed from the index altogether only when \emph{all} B-trees merge in a \emph{single} top-level node. This requires

In summary, by traversing the index's B-trees, the adversary can learn all messages ever sent or received. For a particular deleted message, if this traversal occurs before the original entry and tombstone merge, they can learn the exact content of the message; and if it occurs after the merge, they still learn the \emph{unordered set} of words in a message (since all terms in the tombstone point to position 0). Therefore, in practice, an adversary can leverage FTS4's deletion behavior to look for entries corresponding to a target deleted message, by finding the message's identifier $\messageId$ in the $\messagesTable$ table (by identifying a skip in primary keys for silent features, and rows with NULL messages for visible features) and looking for all B-trees in $\messagesIndexSegdirTable$ and $\messagesIndexSegmentsTable$ that contain the terms for $\messageId$. This can be done efficiently, given that, as explained earlier, B-trees are ordered in chronological order.

\paragraph{Signal.}
Like WhatsApp, Signal maintains a full-text search index across all sent and received messages. They use SQLite's FTS5~\cite{sqlite-fts5} module, which is an upgraded version of FTS4. For our purposes, the main difference is that FT5's deletion


\subsection{Attachment Recovery}\label{subsec:attacks-attachment}
WhatsApp and Signal support five main ways to delete attachment files: the four deletion features mentioned in the preceding section, as well as view-once media (Section~\ref{sec:e2ee-deletion-systematization}). In this section, we show vulnerability against both applications based on \emph{attachment deduplication}, which reveals whether existing attachments were duplicates of deleted ones. So, while the previous section relied on physical remnants of logically-deleted content, here we show how \emph{cross-message application logic} can also be a vector for vulnerabilities, acting as an \emph{implicit} form of physical traces.




% The first two target a particular feature, \emph{view-once media}, where we show how an adversary can learn sensitive information about media files \emph{after} they have been opened. The second vulnerability is general for all deletion features, but requires one assumption about the content of the deleted attachment: that the victim has exchanged at least one more \emph{copy} of it. As we unpack later on, this is a realistic assumption, and has led to attacks in other contexts different from ours. 

\paragraph{File deduplication.} To reduce their storage footprint, both WhatsApp and Signal perform attachment deduplication, that is, they store only one copy of each file. To do so, whenever a new attachment in sent or received, the client compares its checksum against all previously exchanged attachments, storing it in the BLOB store only it is unique. Even if the attachment is a duplicate, however, a row is still inserted to attachment metadata tables $\databaseWhatsApp.\mediaTable$ and $\databaseSignal.\attachmentTable$. Importantly, rows that correspond to subsequent attachment copies reference the metadata in the row of the \emph{original} file, and thus most fields (e.g., location in storage, size, hash, etc.) are identical across rows for all file copies. In particular, columns that store the attachment's \emph{creation time} (i.e., $\mediaKeyTimestampColumn$ in $\databaseWhatsApp.\mediaTable$ and $\uploadTimestamp$ in $\databaseSignal.\attachmentTable$) point to the timestamp of the row in the $\messagesTable$ table corresponding to the message when the file was first exchanged.

More concretely, say a WhatsApp user receives file $F$ (say, an image) at time $t_0$. This adds a binary file to the attachments' BLOB store, as well as new rows in the $\messagesTable$ and $\mediaTable$ tables. In particular, the $\mediaKeyTimestampColumn$ column in the latter has as value $t_0$, equal to the timestamp for the new row in $\messagesTable$. Say then the user receives a second copy of $F$ at a time $t_1 > t_0$. No new file is added to the BLOB store, but new rows are created in the $\messagesTable$ and $\mediaTable$ tables. However, the $\mediaKeyTimestampColumn$ column in the latter has as value $t_0$, indicating deduplication occurred, while the timestamp in $\messagesTable$ is equal to the message's reception time of $t_1$. 
See Figure~\ref{} for an example.

\paragraph{Attachment deletion.} For both WhatsApp and Signal, deleting an attachment-carrying message results in two main changes, which are consistent across all deletion features. First, its corresponding row in $\messagesTable$ is modified accordingly, as explained in Section~\ref{subsec:attacks-text}: either deleted entirely for silent features, or its text cell nullified for visible features. Second, its corresponding row in $\databaseWhatsApp.\mediaTable$ and $\databaseSignal.\attachmentTable$ is deleted entirely (irrespective of the visibility of the feature). Critically, however, if the first instance of a duplicate attachment is deleted, subsequent copies in $\databaseWhatsApp.\mediaTable$ and $\databaseSignal.\attachmentTable$ \emph{still reference the timestamp of the first (now deleted) copy}. Continuing with our example, say the sender of the first copy of $F$ (sent at time $t_0$) unsends the message at a time $t_2$. While this deletes a row in $\databaseWhatsApp.\mediaTable$ (in WhatsApp) and $\databaseSignal.\attachmentTable$ (in Signal), the row corresponding to the second copy (sent at time $t_1$) still has $t_0$ as its timestamp.

As a result, it follows that attachment deduplication reveals the content of deleted attachment files \emph{for which there is at least one future copy of the file}. If so, by inspecting the $\databaseWhatsApp.\mediaTable$ and $\databaseSignal.\attachmentTable$ tables, an adversary identify that existing attachments reference the timestamp of a deleted message, thus learning that it must have been deduplicated with the deleted file. In our example, the second row for $F$ was received at $t_1$ but references timestamp $t_0$ of deleted message, indicating that its content must have been $F$. (In fact, for WhatsApp, if there are three or more copies, we can leverage additional metadata to recover \emph{any} copy of a deleted attachment, irrespective of it being the first one; 
see Appendix~\ref{} for details.) 

Unlike our findings from Section~\ref{subsec:attacks-text}, this vulnerability does not target attachment files generically, depending instead on the aforementioned preconditions of duplicate content. Nonetheless, security should not depend on uniqueness of messages. Indeed, file deduplication has led to vulnerabilities in numerous settings~\cite{fabrega2024exploiting,truong2025breaking,alexeev2025chunking,ritzdorf2016information,harnik2010side}, and thus is a realistic threat vector. This is particularly pressing in the context of messaging applications, where it is common to send duplicate images, even supported explicitly in the form of message forwarding.


\paragraph{Additional vulnerabilities on Signal.} In Signal, file deduplication leads to two additional vulnerabilities, which, surprisingly, allow for recovery of certain \emph{text} messages. Due to space constraints, we discuss these at a high-level here, deferring details to Appendix~\ref{}. 

When long messages are exchanged in Signal, the client displays only a short preview in the chat window, followed by a clickable ``Read more''  label displays the full message in a larger pop-up window. (The exact number of characters the preview varies, depending on the user's device type and other settings such as font size, but a typical cutoff point is around 1000 characters of text.) However, surprisingly, this clickable message preview is encoded and stored as an \emph{image file} in $\databaseSignal$. So, like any other attachment file, future copies of it are deduplicated, and thus our previous vulnerability against duplicate files is directly applicable to (previous of) long messages. So, if there are future copies of (long) messages that are deleted (e.g., a forward of it), an adversary will be able to recover it.

Similarly, the second vulnerability is based on the fact that messages that contain \emph{URLs} are also treated as containing attachment files within $\databaseSignal$, due to the fact that the client automatically ``unfurls'' URLs display an image preview of the website in the chat window~\cite{signal-link-previews}. (Link previews can be disabled in the user settings~\cite{signal-link-previews}, but their default behavior is on.) This image preview is stored as an attachment file within $\databaseSignal$ and thus is subject to the same deduplication vulnerability. So, if there are future copies of a deleted message that contains a URL, an adversary will be able to recover it.

\paragraph{Additional vulnerability on WhatsApp.} Besides the deduplication attacks mentioned thus far, our security analysis uncovered two additional vulnerabilities on WhatsApp, targetting a different mechanism. Unlike deduplication, these attack vectors are more bespoke to WhatsApp's implementation, instead of higher-level design patterns. Furthermore, they are specific to a particular deletion feature, \emph{view-once media}, as opposed to general for all features.

Recall from Section~\ref{sec:e2ee-deletion-systematization} that view-once media allows sending media files that disappear after they have been opened once by the recipient. 
This feature is inherently rooted in exchange of sensitive media, and thus has a number of additional protections built in, e.g., prevents screenshots, forwards, saves, copies, etc~\cite{whatsapp-view-once-media}. 

Inside the application state, as usual, sending or receiving view-once media creates a new row in the $\messagesTable$ and $\mediaTable$ tables of $\databaseWhatsApp$. Among other metadata, the latter stores a \emph{checksum} of the file under its $\fileHashColumn$ column, which is the Base64-encoded SHA256 of the media file. Furthermore, view-once files additionally create a new row in the $\mediaThumbnailTable$ table, which maps message IDs to ``thumbnails'', which are compressed representations of the original file; 
see Figure~\ref{} for an example.

Opening a view-once file (which should render it unrecoverable) leads to two notable changes within $\databaseWhatsApp$, which give rise to two distinct security issues. First, the corresponding row in the $\mediaTable$ table is updated to erase various cells. However, notably the $\fileHashColumn$ column is preserved, and so the adversary is able to recover the \emph{SHA256 checksum of the view-once file}. A file's checksum is highly sensitive information, as it serves as an ``oracle'' to test whether known candidate files match the view-once file, by comparing their hashes with the value in the $\fileHashColumn$ column. So, this allows an adversary to perform \emph{fingerprinting attacks}~\cite{ritzdorf2016information}, i.e., to identify the deleted file from a set of known files.

The second notable issue is that the view-once file's row in the $\mediaThumbnailTable$ table is also preserved, thereby leaking the thumbnail of the view-once file. As such, an adversary can recover a compressed version of the deleted file. While a security issue in isolation, this can also compose with the previous vulnerability, allowing for efficient filtering of candidate files based on the compressed file.

\subsection{Message Reaction Recovery}

\andres{OLD below.}

\paragraph{Vulnerability \#1: write-ahead logging.} The first security issue stems from the \emph{journal mode} used for the SQLite database $\databaseWhatsApp$. In general, a database's journal mode is the mechanism by which new transactions are committed to the database. WhatsApp uses SQLite's \emph{write-ahead logging (WAL)} mode, whereby new changes are first stored in a \emph{second} file, until enough changes accumulate at which point they are written to the main database~\cite{sqlite-journal-mode,sqlite-wal}.

As a result of WAL logging, there is a short window of time in which $\databaseWhatsApp$ is out of sync: any transactions that arise from deleted messages are in the WAL file, but not in $\databaseWhatsApp$. As a result, inspecting $\databaseWhatsApp$ immediately after a deletion feature is triggered allows for. 



\paragraph{Vulnerability \#5: message reactions.}

\paragraph{Vulnerability \#6: pinned messages.} Lastly, we conclude with the simplest of our findings.

This thus reveals to an adversary which messages used to be pinned. This could allow identifying potentially noteworthy or important messages.




