\section{Formal Treatment of E2EE Message Deletion}\label{sec:security-definition}
Our findings from the preceding section show that, like in many other areas, deletion security is a challenge in the E2EE setting. Therefore, in this section, we introduce a formal security model for deletion security of E2EE application clients. We provide syntax that captures the components of clients relevant to deletion, and define security notions that formalizes our threat model and privacy goals. Then, we exercise our model by casting our vulnerabilities from Section~\ref{sec:attacks} as attacks against our security model.

\paragraph{Primitive: deletion client.} We start by formalizing the primitive upon which to base our security definitions. Since our focus is deletion security in end-user devices, our primitive will be \emph{client-centric}, ignoring application servers and other external system components. In particular, we use the client syntax of~\cite{fabrega2025mitigating} as a starting point, which we modify and extend to adapt to our setting. Like~\cite{fabrega2025mitigating}, we narrow our focus to the core components of clients that play a role in our threat model: sending and deleting messages. Formally, we define an \emph{E2EE client} $\client$ as a triple of stateful algorithms $\client = (\Init, \Send, \Delete)$ with the following syntax:

\begin{newitemize}
    \item $\appState_0 \getsr \client.\Init()$: Initializes the client's state to a default starting state $\appState_0$ from a state space\footnote{All spaces in our primitive definition are subsets of $\{0, 1\}^*$.} $\appStateUniverse$.
    \item $\appState_{i+1} \getsr \client.\Send(\appState_i, \userId_s, \{\userId_{r, 1}, \userId_{r, 2}, ...\}, \messageId, M)$: 
    Processes sending message $M$ from message space $\messageUniverse$, with (unique) message identifier $\messageId \in \messageIdUniverse$, from sender $\userId_s$ to set of recipients $\{\userId_{r, 1}, \userId_{r, 2}, ...\}$. All $\userId$ are unique user identifiers from space $\userIdUniverse$, and we require that the ID for the client owner, denoted by $\userId_\client$, is either the sender or in the recipient set. Returns the updated state $\appState_{i+1} \in \appStateUniverse$ as a result.
    \item $\appState_{i+1} \getsr \client.\Delete(\appState_i, \messageId, \deletionMetadata)$: Processes deletion of message with identifier $\messageId \in \messageIdUniverse$, in accordance with some (optional) deletion metadata $\deletionMetadata$ from space $\deletionMetadataUniverse$.
\end{newitemize}
We make a few brief remarks to provide more context on our primitive; see Appendix~\ref{} ƒor details, including a full example of real-world clients map to our syntax. 
State $\appState$ corresponds to the (physical) application state, such a serialized database in storage. Identifiers $\userId$ and $\messageId$ are application-specific representations of users and messages in $\appState$, such as phone numbers for the latter and counters for the latter. Following other works (e.g.,~\cite{e2ee-cloud-storage,fabrega2025mitigating}), we model ``messages'' as encoding both their content and relevant metadata (e.g., timestamps) in $M$. Furthermore, we use $\Send$ as a single abstraction over all the ways in which users may interact (e.g., sending texts, attachments, etc.), which are disambiguated based on information encoded in $M$. Similarly, $\Delete$ is an abstraction over all deletion features. We encode feature-specific metadata in $\deletionMetadata$, such as the feature that triggered deletion, the time of deletion, etc. Finally, notice that $\Send$ and $\Delete$ are \emph{randomized} algorithms. This is because clients may process new messages or deletions non-deterministically, e.g., if parts of $\appState$ are encrypted.

We define a \emph{transcript} $\transcript$ to be an ordered sequence of $\Send$ and $\Delete$ operations together with their non-state inputs. That is, each entry in $T$ is a tuple of the form $(\Send, \userId_s, \{\userId_1, \userId_2, ...\}, \messageId, M)$ or $(\Delete, \messageId, \deletionMetadata)$. Notice that a transcript induces a (probability distribution over) final state, by calling $\client.\Init()$ followed by running each operation in $\transcript$ in order. We denote this via algorithm $\appState \getsr \client.\Run(\transcript)$, defined formally in Figure~\ref{}.

As discussed in earlier sections, our security notions will also require modelling a user's ``logical'' application content, that is, the set of exchanged messages displayed in the client's UI (non-deleted messages, and placeholder strings for visible deletion features). We model this as a mapping $\content : \mathcal{P}(\userIdUniverse) \to (\messageUniverse \cup \{\bot\})^*$ from sets of user IDs (representing conversations) to tuples of message and a special deletion symbol $\bot$ (representing exchanged messages therein). For example, ...

Like $\appState$, the mapping $\content$ is implicitly defined by $\Send$ and $\Delete$ operations, since these represent the flow of messages. In particular, a $\Send$ operation appends $M$ to the list under key (i.e., conversation) of user IDs $\{\userId_s, \userId_{r, 1}, \userId_{r, 2}, ...\}$. Then, a $\Delete$ operation can modify $\content$ in one of two ways, depending on the type of deletion feature that is executed: it removes $M$ altogether for silent features, or replaces it with $\bot$ for visible features (since, recall, the former type removes messages from the UI, while the latter type leaves a placeholder string).
We denote by $\content \gets \client.\getContent(\transcript)$ the logical content induced by the operations in $T$; see Figure~\ref{} for its formal definition.
Notice that $\client.\getContent$ is deterministic, while $\client.\Run$ is a randomized: a sequence of operations defines a single logical state, but defines a probability distribution over physical states (since, recall, $\Send$ and $\Delete$ are randomized algorithms).

\paragraph{Defining secure client-side deletion.} Having introduced our primitive syntax, we now turn to defining security. Recall from Section~\ref{sec:threat-model} that, at a high-level, our goal is for physical device access to reveal nothing more about messages than can be learned from the user's logical state (i.e., what is displayed in their user interface). We can formally cast this goal as a form of \emph{history independence}~\cite{micciancio1997oblivious,naor2001anti}, that is, requiring that a client's storage representation does not reveal information about the history of operations (in particular, deletions). History independence is natural motivation for our setting, as it bears close relationship with the general problem of deletion privacy; indeed, it has been the basis for other deletion definitions in contexts different from ours~\cite{cohen2023control}.

Let us proceed to make this concrete. First, we say that transcripts $T_0$ and $T_1$ are \emph{equivalent} if they induce the same logical application content, i.e., if they lead to the same $\content$. For example, ... Formally:
\begin{definition}\label{def:transcript-equivalence}
    Let $\client$ be an E2EE client, and $T_0$ and $T_1$ be two transcripts. Let $\content_0 \gets \client.\getContent(T_0)$ and $\content_1 \gets \client.\getContent(T_1)$. Transcripts $T_0$ and $T_1$ are \emph{equivalent} if $\content_0 = \content_1$.
\end{definition}
Moving forward, we will use $T_0 \equiv T_1$ as a shorthand to denote that $T_0$ and $T_1$ are equivalent. \andres{Note on how this depends on C, which we leave implicit in our notation?} With this notion, we can then frame our security goals as requiring that \emph{any pair of equivalent transcripts leads to indistinguishable application states}.
We term this security guarantee \emph{deletion indistinguishability} ($\SecDef$), which we define formally with the game shown in Figure~\ref{fig:ind-del-def}. It is based on standard left-or-right (LOR) indistinguishability security notions~\cite{}, but adapted to our deletion security goals. In more detail, $\SecDef$ provides two oracles to the adversary $\advA$: the first is an honest oracle, which allows $\advA$ to input a transcript and obtain the resulting application state; and the second is a \emph{challenge} oracle, which allows $\advA$ to input \emph{pairs} of transcripts and obtain the resulting state for one of them, as determined by a randomly-chosen challenge bit, subject to the constraint that both challenge transcripts are equivalent. So, conceptually, $\SecDef$ can be though of as a blend of LOR indistinguishability and history independence. We define the advantage of a $\SecDef$ adversary $\advA$ over E2EE client $\client$ as 
\[\Advantage{\SecDefShort}{\client}(\advA) = \big|2 \cdot \Prob{\SecDef^\advA_\client} - 1\big|\]

\input{Figures/ind-del-def}

Intuitively, $\SecDef$ models an adversary that can generate arbitrary sequences of $\Send$ and $\Delete$ operations, and observe the resulting physical device state. Then, to break security, they must generate a sequence that leads to identical logical content (i.e., messages displayed in the UI) but different physical application states; in other words, as a result of deletions, the application state is revealing more information that the logical state, thus formally capturing our target threat model and security goals.


Notice that $\SecDef$ operates over transcripts as opposed to individual operations (that is, $\advA$ does not inspect the state at every $\Send$ or $\Delete$). \andres{Stretch goal}....

\paragraph{IND-DEL via transcript distance.} While $\SecDef$ characterizes our desired security guarantees, proving security in some settings may be complex. So, we now distill a simpler, higher-level property of transcripts, which we show is equivalent to $\SecDef$, thus providing a simpler basis to reason about deletion security. At the core of our simplification is the insight that, intuitively, $\SecDef$ can be thought of as measuring the \emph{statistical distance} between the distribution of application states induced by equivalent transcripts. So, we can bound $\Advantage{\SecDefShort}{}$ by the maximum statistical distance over all pairs of transcripts. We refer to this property as \emph{state consistency}, reflecting the fact that application states are consistent for equal logical states. We formally define it as follows:
\begin{definition}\label{def:state-consistency}
    E2EE client $\client$ is \emph{$\epsilon$-state consistent} if, for any transcripts $\transcript_0$ and $\transcript_1$ such that $\transcript_0 \equiv \transcript_1$, it follows that 
    \[\delta\big(\client.\Run(\transcript_0), \client.\Run(\transcript_1)\big) \leq \epsilon\]
    where $\delta$ denotes the total variation distance between both random variables.
\end{definition}

In other words, for any two equivalent transcripts, the probability distribution over application states is at most $\epsilon$ apart. We then capture the relationship between $\SecDef$ and state consistency in the following simple theorem:
\begin{theorem}\label{thm:ind-del-state-consistency}
    Let $\client$ be an E2EE client that is $\epsilon$-state consistent, and $\advA$ be an $\SecDef$ adversary that makes $q$ challenge oracle queries. Then, it follows that
    \[\Advantage{\SecDefShort}{\client}(\advA) \leq q \cdot \epsilon\]
\end{theorem}
We show the proof of this theorem in Appendix~\ref{}. 
It follows from the fact that total variation distance is an upper bound on the advantage of any distinguisher that observes a sample from either distribution~\cite{maurer2007indistinguishability}, and so state consistency yields an upper bound on the distinguishing advantage of $\advA$ in every challenge oracle query. 
A standard hybrid argument then leads to the claimed bound.


\paragraph{Leakage-based generalization.} Thus far, our $\SecDef$ definition represents a strong guarantee: application states must be \emph{identically distributed} for equivalent operation transcripts, allowing not a single bit of information leakage. However, in practice, a small amount of well-defined leakage may be desirable (or even necessary). For example, consider an $\appState$ that uses an increasing counter as the primary key for a database table where all messages are stored. In this case, satisfying $\SecDef$ would require shifting all primary keys by 1 whenever a message is deleted. Otherwise, the fact that message $i$ was deleted can be determined from the fact that primary keys jump from $i-1$ to $i+1$. Adjusting all database rows after a deletion would lead to high runtime overheads, even though the only thing learned by the adversary is the total number of deleted messages, but nothing about their content.

We thus relax our definition to allow for a controlled amount of information leakage. Drawing from other lines of work with leakage-based definitions (e.g., searchable encryption~\cite{}),
we parametrize our security model with a \emph{leakage function} $\leakage$, which ``filters'' adversarial challenge queries to model some information that is allowed to leak. Concretely, we define $\SecDef_{\client}^{\advA, \leakage}$ as identical to $\SecDef_{\client}^{\advA}$, but additionally verifying that $\leakage(\transcript_0) = \leakage(\transcript_1)$ at the start of the $\ChRunOracle$ oracle, returning $\bot$ if not. Similarly, we generalize Definition~\ref{def:state-consistency} to now require that $\transcript_0 \equiv \transcript_1$ \emph{and} $\leakage(\transcript_0) = \leakage(\transcript_1)$, and say that $\client$ is $\epsilon$-state consistent \emph{with respect to leakage function $\leakage$} if it holds. Notice that Theorem~\ref{thm:ind-del-state-consistency} extends directly to this generalized setting, obtaining an identical bound for $\Advantage{\SecDefShort}{\client, \leakage}(\advA)$ from $\epsilon$-state independence with respect to $\leakage$.

By normalizing for identical leakage functions in challenge transcripts, $\SecDef$ now models the fact that we allow leaking the output of $\leakage$ to $\advA$. In particular, our earlier treatment can be cast a special case of this broader framing, i.e., when $\leakage$ is the identity function. As a second example, we can frame the example from earlier of skips in primary keys in terms of a leakage function that reveals the total number of $\Delete$ operations, i.e., \[\leakageDelNum(\transcript) = \big|\{\textsf{op} \in \transcript\ |\ \textsf{op}[0]=\Delete\}\big|.\]

While leakage-based definitions have the benefit of being flexible, allowing for any instantiation of $\leakage$, they naturally raise the question of \emph{what is a good leakage regime to target}. For instance, consider $\leakageDelNum$ from above. What sensitive information can an attacker learn from the number of deleted messages, if any? This tension is inherent to any leakage-based security model. So, to address this, we follow the approach of~\cite{tyagi2018burnbox}, and complement $\SecDef$ with a property-based security definition that captures the more specific adversarial goal of indistinguishability of deleted messages. \andres{Intuition of game here. Say it provides a basis upon which to test leakage functions. In particular, delN meets it. However, exploring the space of leakage functions (comparing them, analyzing their security) is left as future work, akin to how it is in other fields.}

\paragraph{Analysis of vulnerabilities.} Our security model provides formal foundations to reason about deletion security, both in terms of analyzing vulnerabilities and evaluating proposed defenses. To exemplify this, we now cast our findings from Section~\ref{sec:attacks} as attacks against our definitions.

\andres{Somewhere say that this allows us to formalize analyze attacks, as well as evaluating the security of defenses. Indeed, looking ahead we do this in section 7 with our proposed mitigations.}